---
title: "Assignment3"
author: "Nanna Mathiasen, Simon Hansen, Alberte Seeberg, Esther Dyngby"
date: "3/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Multivariate linear models

In this assignment you will investigate the relation between different aspects of IQ and symptom severity in ASD (as measured by ADOS). The assignment is meant to make you practice on linear models, multiple predictors and interactions.

The data you will use is the outcome of the psychological testing of the children you analyzed in methods 3.
Data: https://www.dropbox.com/s/hom2qnmp9hyfwmu/Assignment3Data.csv?dl=0
The data consists of ChildID, gender, age, diagnosis (ASD), symptom severity (ADOS), several aspects of IQ (NonVerbal/Perceptual,Verbal,Social) and the visit at which the tests were administered. The questions should be answered by focusing on visit 1.

The questions you will have to answer are:

1. Assess the relation between symptom severity and IQ (focus on visit 1 and children with ASD) and report the model, a plot of the model, a couple of lines describing the quality of the model and interpreting the results. P.S. Should you scale? P.P.S. Don't forget to motivate your priors. P.P.P.S. At least one plot for results and a plot for quality of each model (here and in the next questions) would be appreciated.
1.1. Verbal IQ and ADOS
1.2. Non Verbal IQ and ADOS
1.3. Social IQ and ADOS

```{r setup, include=FALSE}
data = read.csv("Assignment3Data.csv")
data = subset(data, Visit == 1)
data = subset(data, ASD == 1)
library(rethinking)


#Scaling
data$NonVerbalIQ = scale(data$NonVerbalIQ)
data$VerbalIQ = scale(data$VerbalIQ)
data$SocialIQ = scale(data$SocialIQ)

plot(data$ADOS ~ data$VerbalIQ)
#Making the model for verbal IQ
verbalIQ <- map(alist(ADOS~dnorm(mu, sigma), 
                  mu <- a + b*VerbalIQ, 
                  a ~ dnorm(13,10),
                  b~ dnorm(0,5),
                  sigma ~ dunif(0,5)), data = data)
#What are the values of the model
precis(verbalIQ, corr = TRUE)

#Plotting the values
# Adjusting the scale, we want values from -10 to 15. "by" makes a datapoint for each 0.1. 
ScaleVIQ.seq <- seq( from=-10 , to=15 , by=0.1 )
# Simulating 1000 data points for each point on the scale
mu <- link(verbalIQ, data=data.frame(VerbalIQ=ScaleVIQ.seq) )
str(mu)
#Calculate mean and HPDI for each data point
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )

#Simulate data points from the model in the sequence just made.
sim.ADOS <- sim(verbalIQ , data=list(VerbalIQ=ScaleVIQ.seq) )
#Calculating HPDI for each data point in the model so we can investigate the model and make the plot
ADOS.HPDI <- apply( sim.ADOS , 2 , HPDI , prob=0.89 )

# plot raw data
plot(ADOS ~ VerbalIQ , data , col=col.alpha(rangi2,0.5) )
# draw MAP line, there is a point for each 0.1 so it becomes a line 
lines( ScaleVIQ.seq , mu.mean )
# draw HPDI region for line, there are two HPDI values making a shade that shows the uncertainty of the line/model
shade(mu.HPDI , ScaleVIQ.seq)
# draw HPDI region for simulated ADOS, the percentile interval. The uncertainty of the variables in the model
shade(ADOS.HPDI , ScaleVIQ.seq)



### SOCIAL IQ
plot(data$ADOS ~ data$SocialIQ)

#Making the model for social IQ
socialIQ <- map(alist(ADOS~dnorm(mu, sigma), 
                  mu <- a + b*SocialIQ, 
                  a ~ dnorm(13,10),
                  b~ dnorm(0,5),
                  sigma ~ dunif(0,5)), data = data)
#What are the values of the model
precis(socialIQ, corr = TRUE)

#Plotting the values
# Adjusting the scale, we want values from -10 to 15. by makes a datapoint for each 0.1. 
ScaleVIQ.seq <- seq( from=-20 , to=30 , by=0.1 )
# Simulating 1000 data points for each point on the scale
mu <- link(socialIQ, data=data.frame(SocialIQ=ScaleVIQ.seq) )
str(mu)
#Calculate mean and HPDI for each data point
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )

#Simulate data points from the model in the sequence just made.
sim.ADOS <- sim(socialIQ , data=list(SocialIQ=ScaleVIQ.seq) )
#Calculating HPDI for each data point in the model so we can investigate the model and make the plot
ADOS.HPDI <- apply( sim.ADOS , 2 , HPDI , prob=0.89 )

# plot raw data
plot(ADOS ~ SocialIQ , data , col=col.alpha(rangi2,0.5) )
# draw MAP line, there is a point for each 0.1 so it becomes a line 
lines( ScaleVIQ.seq , mu.mean )
# draw HPDI region for line, there are two HPDI values making a shade that shows the uncertainty of the line
shade(mu.HPDI , ScaleVIQ.seq)
# draw HPDI region for simulated ADOS, the percentile interval. 
shade(ADOS.HPDI , ScaleVIQ.seq)




### Nonverbal IQ
plot(data$ADOS ~ data$NonVerbalIQ)

#Making the model for social IQ
nonverbalIQ <- map(alist(ADOS~dnorm(mu, sigma), 
                  mu <- a + b*NonVerbalIQ, 
                  a ~ dnorm(13,10),
                  b~ dnorm(0,5),
                  sigma ~ dunif(0,5)), data = data)
#What are the values of the model
precis(nonverbalIQ, corr = TRUE)

#Plotting the values
# Adjusting the scale, we want values from -10 to 15. by makes a datapoint for each 0.1. 
ScaleVIQ.seq <- seq( from=-20 , to=30 , by=0.1 )
# Simulating 1000 data points for each point on the scale
mu <- link(nonverbalIQ, data=data.frame(NonVerbalIQ=ScaleVIQ.seq) )
str(mu)
#Calculate mean and HPDI for each data point
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )

#Simulate data points from the model in the sequence just made.
sim.ADOS <- sim(nonverbalIQ , data=list(NonVerbalIQ=ScaleVIQ.seq) )
#Calculating HPDI for each data point in the model so we can investigate the model and make the plot
ADOS.HPDI <- apply( sim.ADOS , 2 , HPDI , prob=0.89 )

# plot raw data
plot(ADOS ~ NonVerbalIQ , data , col=col.alpha(rangi2,0.5) )
# draw MAP line, there is a point for each 0.1 so it becomes a line 
lines( ScaleVIQ.seq , mu.mean )
# draw HPDI region for line, there are two HPDI values making a shade that shows the uncertainty of the line
shade(mu.HPDI , ScaleVIQ.seq)
# draw HPDI region for simulated ADOS, the percentile interval. 
shade(ADOS.HPDI , ScaleVIQ.seq)
```


2. Do the different aspects of IQ account for different portions of the variance in ADOS? 
2.1. Does it make sense to have all IQ measures in the same model? First write a few lines answering the question and motivating your answer, including a discussion as to what happens when you put all of them in the same model. Then build a model following your answer. If your answer is "no", you are not free, you still have to answer: are there alternative ways of answering the question?
2.2. Build the model, assess its quality, write a few lines interpreting the results.

```{r setup, include=FALSE}
library(rethinking)
library(cowplot)
library(ggplot2)
#Loading data
data2 = read.csv("Assignment3Data.csv")
data2 = subset(data2, Visit == 1)
data2 = subset(data2, ASD == 1)

### Correlation test
cor.test(data2$VerbalIQ, data2$SocialIQ)
cor.test(data2$VerbalIQ, data2$NonVerbalIQ)
cor.test(data2$NonVerbalIQ, data2$SocialIQ)


#Scaling the data
data3 = read.csv("Assignment3Data.csv")
data3 = subset(data3, Visit == 1)
data3 = subset(data3, ASD == 1)

data3$NonVerbalIQ = scale(data3$NonVerbalIQ)
data3$VerbalIQ = scale(data3$VerbalIQ)
data3$SocialIQ = scale(data3$SocialIQ)

#To investigate the unique contribution of each variable we create new models
#Social model
m_S <- map(alist(SocialIQ~dnorm(mu, sigma), 
                  mu <- a + b_V*VerbalIQ + b_NV*NonVerbalIQ, 
                  a ~ dnorm(0,5),
                  b_V ~ dnorm(0,5),
                  b_NV ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data3)
mu_S <- coef(m_S)['a'] + coef(m_S)['b_V']*data3$VerbalIQ + coef(m_S)['b_NV']*data3$NonVerbalIQ
#compute residual for each state
m.resid_S <- data3$SocialIQ - mu_S


plot_S <- ggplot(data3, aes(x = m.resid_S, y = ADOS)) + geom_smooth(method = "lm") + geom_point() + xlab("SocialIQ residuals")
plot_S

#Verbal model
m_V <- map(alist(VerbalIQ~dnorm(mu, sigma), 
                  mu <- a + b_S*SocialIQ + b_NV*NonVerbalIQ, 
                  a ~ dnorm(0,5),
                  b_S ~ dnorm(0,5),
                  b_NV ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data3)
mu_V <- coef(m_V)['a'] + coef(m_V)['b_S']*data3$VerbalIQ + coef(m_V)['b_NV']*data3$NonVerbalIQ
#compute residual for each state
m.resid_V <- data3$VerbalIQ - mu_V


plot_V <- ggplot(data3, aes(x = m.resid_V, y = ADOS)) + geom_smooth(method = "lm") + geom_point() + xlab("VerbalIQ residuals")
plot_V

#Nonverbal model
m_NV <- map(alist(NonVerbalIQ~dnorm(mu, sigma), 
                  mu <- a + b_V*VerbalIQ + b_S*SocialIQ, 
                  a ~ dnorm(0,5),
                  b_V ~ dnorm(0,5),
                  b_S ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data3)
mu_NV <- coef(m_NV)['a'] + coef(m_NV)['b_V']*data3$VerbalIQ + coef(m_NV)['b_S']*data3$SocialIQ
#compute residual for each state
m.resid_NV <- data3$NonVerbalIQ - mu_NV


plot_NV <- ggplot(data3, aes(x = m.resid_NV, y = ADOS)) + geom_smooth(method = "lm") + geom_point() + xlab("NonverbalIQ residuals")
plot_NV

plot_grid(plot_S, plot_V, plot_NV)

# To add the residuals found above to the data 

data3$NVIQ_res = m.resid_NV
data3$VIQ_res = m.resid_V
data3$SIQ_res = m.resid_S

# Creating a model that predicts ADOS from the residuals 
m_NonVerbal_res=map(
  alist(
    ADOS ~ dnorm( mu , sigma ),
    mu <- Intercept +
        b_NVIQ*NVIQ_res,
    Intercept ~ dnorm(0,5),
    b_NVIQ ~ dnorm(0,5),
    sigma ~ dunif(0,10)
), data = data3)

precis(m_NonVerbal_res)

m_Verbal_res=map(
  alist(
    ADOS ~ dnorm( mu , sigma ),
    mu <- Intercept +
        b_VIQ*VIQ_res,
    Intercept ~ dnorm(0,5),
    b_VIQ ~ dnorm(0,5),
    sigma ~ dunif(0,10)
), data = data3)

precis(m_Verbal_res)

m_Social_res=map(
  alist(
    ADOS ~ dnorm( mu , sigma ),
    mu <- Intercept +
        b_SIQ*SIQ_res,
    Intercept ~ dnorm(0,5),
    b_SIQ ~ dnorm(0,5),
    sigma ~ dunif(0,10)
), data = data3)

precis(m_Social_res)


# Building a model with the relevant IQ measures in (all IQ measures). 
m1 <- map(alist(ADOS~dnorm(mu, sigma), 
                  mu <- a + b_S*SocialIQ + b_V*VerbalIQ + b_NV*NonVerbalIQ, 
                  a ~ dnorm(0,5),
                  b_S ~ dnorm(0,5),
                  b_V ~ dnorm(0,5),
                  b_NV ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data3)
#What are the values of the model
precis(m1)
plot(precis(m1))


#Quality check of the model
mu <- link( m1)
# calculating mean and HPDI for each data point
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI )
# simulate observations
ADOS.sim <- sim( m1 , n=1e4 )
ADOS.HPDI <- apply( ADOS.sim , 2 , HPDI )

#plotting
plot( mu.mean ~ data3$ADOS , col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
abline( a=0 , b=1 , lty=2 )
for ( i in 1:nrow(data3) )
lines( rep(data3$ADOS[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )

```

3. Let's now include also the TD children. Does it make sense to ask whether IQ and ADOS are related? Motivate your answer. In any case, if you wanted to build a model to answer that question, which model would you build? Run the model, assess its quality, write a few lines interpreting the results.

```{r setup, include=FALSE}
data4 = read.csv("Assignment3Data.csv")
data4 = subset(data4, Visit == 1)

#Scaling
data4$NonVerbalIQ = scale(data4$NonVerbalIQ)
data4$VerbalIQ = scale(data4$VerbalIQ)
data4$SocialIQ = scale(data4$SocialIQ)

data4 = na.omit(data4)

m4 <- map(alist(ADOS~dnorm(mu, sigma), 
                  mu <- a + b_S*SocialIQ + b_V*VerbalIQ + b_NV*NonVerbalIQ, 
                  a ~ dnorm(0,5),
                  b_S ~ dnorm(0,5),
                  b_V ~ dnorm(0,5),
                  b_NV ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data4)
#What are the values of the model
precis(m4)
plot(precis(m4))

#Quality check of the model
mu <- link( m4)
# calculating mean and HPDI for each data point
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI )
# simulate observations
ADOS.sim <- sim( m4 , n=1e4 )
ADOS.HPDI <- apply( ADOS.sim , 2 , HPDI )

#Making ASD a factor to plot it
data4$ASD = as.factor(data4$ASD)

library(RColorBrewer)
#all palette available from RColorBrewer
display.brewer.all()
#we will select the first 2 colors in the Set1 palette
cols<-brewer.pal(n=2,name="Set1")
#cols contain the names of four different colors
#create a color vector corresponding to levels in the T1 variable in dat
cols_ASD<-cols[data4$ASD]

#plotting
plot( mu.mean ~ data4$ADOS , data4, col=cols_ASD , pch=16, ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
abline( a=0 , b=1 , lty=2 )
for ( i in 1:nrow(data4) )
lines( rep(data4$ADOS[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )


#Adding ASD as predictor
data4$ASD = as.integer(data4$ASD)
glimmer(ADOS ~ SocialIQ + VerbalIQ + NonVerbalIQ + ASD, data4)
m5 <- map(alist(ADOS~dnorm(mu, sigma), 
                  mu <- a + b_S*SocialIQ + b_V*VerbalIQ + b_NV*NonVerbalIQ + b_ASD*ASD, 
                  a ~ dnorm(0,10),
                  b_S ~ dnorm(0,1),
                  b_V ~ dnorm(0,1),
                  b_NV ~ dnorm(0,1),
                  b_ASD ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data4)
#What are the values of the model
precis(m5)
plot(precis(m5))


```

4. Let's discuss contents:
4.1. You have three scores for IQ, do they show shared variance? Is that the same in TD and ASD? What does that tell us about IQ?
4.2. You have explored the relation between IQ and ADOS. How do you explain that relation from a cognitive perspective? N.B. You can present alternative hypotheses.

```{r setup, include=FALSE}
#Loading data, only TDs
data5 = read.csv("Assignment3Data.csv")
data5 = subset(data5, Visit == 1)
data5 = subset(data5, ASD == 0)
data5 = na.omit(data5)

#Scaling the data
data5$NonVerbalIQ = scale(data5$NonVerbalIQ)
data5$VerbalIQ = scale(data5$VerbalIQ)
data5$SocialIQ = scale(data5$SocialIQ)

# Create residual plots for each IQ
#Social model
m_S_TD <- map(alist(SocialIQ~dnorm(mu, sigma), 
                  mu <- a + b_V*VerbalIQ + b_NV*NonVerbalIQ, 
                  a ~ dnorm(0,5),
                  b_V ~ dnorm(0,5),
                  b_NV ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data5)
mu_S_TD <- coef(m_S_TD)['a'] + coef(m_S_TD)['b_V']*data5$VerbalIQ + coef(m_S_TD)['b_NV']*data5$NonVerbalIQ
#compute residual for each state
m.resid_S_TD <- data5$SocialIQ - mu_S_TD


plot_S_TD <- ggplot(data5, aes(x = m.resid_S_TD, y = ADOS)) + geom_smooth(method = "lm") + geom_point() + xlab("SocialIQ residuals")
plot_S_TD

#Verbal model
m_V_TD <- map(alist(VerbalIQ~dnorm(mu, sigma), 
                  mu <- a + b_S*SocialIQ + b_NV*NonVerbalIQ, 
                  a ~ dnorm(0,5),
                  b_S ~ dnorm(0,5),
                  b_NV ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data5)
mu_V_TD <- coef(m_V_TD)['a'] + coef(m_V_TD)['b_S']*data5$VerbalIQ + coef(m_V_TD)['b_NV']*data5$NonVerbalIQ
#compute residual for each state
m.resid_V_TD <- data5$VerbalIQ - mu_V_TD


plot_V_TD <- ggplot(data5, aes(x = m.resid_V_TD, y = ADOS)) + geom_smooth(method = "lm") + geom_point() + xlab("VerbalIQ residuals")
plot_V_TD

#Nonverbal model
m_NV_TD <- map(alist(NonVerbalIQ~dnorm(mu, sigma), 
                  mu <- a + b_V*VerbalIQ + b_S*SocialIQ, 
                  a ~ dnorm(0,5),
                  b_V ~ dnorm(0,5),
                  b_S ~ dnorm(0,5),
                  sigma ~ dunif(0,10)), data = data5)
mu_NV_TD <- coef(m_NV_TD)['a'] + coef(m_NV_TD)['b_V']*data5$VerbalIQ + coef(m_NV_TD)['b_S']*data5$SocialIQ
#compute residual for each state
m.resid_NV_TD <- data5$NonVerbalIQ - mu_NV_TD


plot_NV_TD <- ggplot(data5, aes(x = m.resid_NV_TD, y = ADOS)) + geom_smooth(method = "lm") + geom_point() + xlab("NonverbalIQ residuals")
plot_NV_TD

plot_grid(plot_S_TD, plot_V_TD, plot_NV_TD)






```

5. Bonus questions: Including measurement errors. 
5.1. Let's assume that ADOS has a measurement error of 1. How would you model that in any of the previous models? 
5.2. We know that IQ has an estimated measurement error of 2.12. How would you include that? 


